Abstractions for Type-Level Programming

### 2022-03-21 slides.tex
slides

\usepackage{xspace}
\usepackage{mathtools}
\def\A{\normalfont\text{A}}
\def\B{\normalfont\text{B}}
\def\C{\normalfont\text{C}}
\def\D{\normalfont\text{D}}
\def\T{\normalfont\text{T}}
\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}

@@

The regex library uses match types, a new language construct introduced in this thesis.
\end{slide}

\begin{slide}{Compilation times for the type-level join operation}
\begin{center}
\input{figures/join.gex}
\end{center}
\end{slide}

@@

\begin{slide}{Singleton types}
Scala supports a few singleton types:
\begin{itemize}
  \item |x.type|, the type of the variable |x| (since forever)
  \item |42|, the type of the integer literal |42| (since 2016)
  \item |+|, the type of integer addition (since 2020)
\end{itemize}
\pause
Question: How much of the language can we represent in types?
\end{slide}

@@

\begin{slide}{"Precise mode" of type inference: list concatenation example}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}
\begin{overprint}
\onslide<1-2>
\begin{code}
dependent val l1 = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

\onslide<3>
\begin{code}
val l1: List     = "A" :: Nil
dependent val l2 = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { concat(l1, l2) }
\end{code}

\onslide<4>
\begin{code}
dependent val l1 = "A" :: Nil
val l2: List     = "B" :: Nil
dependent val l3 = concat(l1, l2)
l3: { "A" :: l2 }
\end{code}

@@

\begin{slide}{Type evaluation}
During subtyping, we evaluate both sides:
\begin{code}
A <: B  iff  eval(A) <: eval(B)
\end{code}
\pause

Straightforward for if-then-else:
\begin{code}
eval(If[true,  A, B]) = A
eval(If[false, A, B]) = B
\end{code}
\pause

More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
\end{slide}
\begin{slide}{Type evaluation: type tests}
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
\end{itemize}
\end{slide}

@@

|foo(readInt(): { (_: Int) + 1 })|
\end{slide}

\begin{slide}{Generalized singletons: recap}

We proposed a generalization of Scala's singleton types:
\begin{enumerate}
  \item lift a subset of Scala's language constructs to the type level
  \item add a "precise mode" of type inference
  \item evaluate types during subtyping
\end{enumerate}
\end{slide}

% Termination
% Virtual dispatch

@@

\begin{overprint}
\onslide<2-5>
\begin{code}
Elem[String] =:= Char
\end{code}

\onslide<3-5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
\end{code}

\onslide<4-5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{code}

\onslide<5>
\begin{code}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
Elem[Seq[Int]] =:= Elem[Seq[Int]]
\end{code}

\onslide<6>
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
\end{itemize}

\onslide<7>
Match types can be explained in terms of generalized singletons:

@@

\onslide<8>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X <: Singleton](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] @=>@ elem(x.head)
  case x: Any => x
\end{code}

\onslide<9>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X <: Singleton](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x

elem[Any]("foo")
\end{code}
\vspace{-20pt}
\end{overprint}
\end{slide}

\begin{slide}{Formalization}
System FM is an extension of System \Fsub with
\begin{itemize}
  \item pattern matching
  \item opaque classes
  \item match types
\end{itemize}
\pause
Design goals:
\begin{itemize}
  \item explain the essence of match types
  \item give us confidence in our design
  \item be as simple as possible
\end{itemize}
\end{slide}

@@

\begin{slide}{System FM's pattern matching evaluation}
Given a concrete scrutinee, we evaluate match expressions with a few lookups in the inheritance relation:
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/FMEMatch.pdf}
\end{center}
\end{slide}

\begin{slide}{System FM's disjointness relation}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/FMDisjointness.pdf}
\end{center}
\end{slide}

\begin{slide}{System FM's match types evaluation rule (subtyping)}
\includegraphics[width=\textwidth]{figures/FMSMatch.pdf}
\end{slide}

@@

\begin{slide}{Empty types}
Our system does not like empty types.

@@

Intersections require special care.

@@

\begin{slide}{Variance}
At first glance, we can't prove disjointness when variance is involved.

|F[+T]:| \quad |T1 <: T2| \quad implies \quad |F[T1] <: F[T2]|\\
|G[-T]:| \quad |T1 <: T2| \quad implies \quad |G[T1] >: G[T2]|
\pause

@@

\appendix

\begin{slide}{Type-safe regular expressions (simplified)}
\begin{code}
val rational = new Regex("(\\d+)\\.(\\d+)")
rational.unapply("3.1415"): Option[(String, String)]

type GroupsOf[P <: String] =
  Reverse[Loop[P, 0, Length[P], EmptyTuple]]

type Loop[P, Lo, Hi, Acc <: Tuple] =
  Lo match
    case Hi => Acc
    case _ => CharAt[P, Lo] match
      case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
      case '\\' => Loop[P, Lo + 2, Hi, Acc]
      case _ => Loop[P, Lo + 1, Hi, Acc]
\end{code}
\end{slide}

>>>>

\usepackage{ulem}
\usepackage{xspace}
\usepackage{mathtools}
\def\A{\normalfont\text{A}}
\def\B{\normalfont\text{B}}
\def\C{\normalfont\text{C}}
\def\D{\normalfont\text{D}}
\def\T{\normalfont\text{T}}
\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}

@@

The regex library uses technology developed in this thesis.
\end{slide}

\begin{slide}{Compilation times for the type-level join operation}
\begin{center}
\input{figures/join-slides.gex}
\end{center}
\end{slide}

@@

\begin{slide}{Singleton types}
\begin{block}{Definition}
A singleton type is a type that contains exactly one value.
\end{block}

Scala has a few of those:
\begin{itemize}
  \item |x.type|, the type of the variable |x| (since forever)
  \item |42|, the type of the integer literal |42| (since 2016)
  \item |+|, the type of integer addition (since 2020)
\end{itemize}
\pause
Question: How much more of the language can we represent as singleton types?
\end{slide}

@@

\begin{slide}{"Precise mode" of type inference: list concatenation example}
\begin{code}
dependent def concat(xs: List, ys: List) <: List =
  xs match
    case x :: xs => x :: concat(xs, ys)
    case Nil => ys
\end{code}

\begin{overprint}
\onslide<1>
\begin{code}
dependent val l1 @=@ "A" :: Nil
dependent val l2 @=@ "B" :: Nil
dependent val l3 @=@ concat(l1, l2)
l3: { "A" :: "B" :: Nil }
\end{code}

\onslide<2>
\begin{code}
val l1: List     @=@ "A" :: Nil
dependent val l2 @=@ "B" :: Nil
dependent val l3 @=@ concat(l1, l2)
l3: { concat(l1, l2) }
\end{code}

\onslide<3>
\begin{code}
dependent val l1 @=@ "A" :: Nil
val l2: List     @=@ "B" :: Nil
dependent val l3 @=@ concat(l1, l2)
l3: { "A" :: l2 }
\end{code}

@@

\begin{slide}{Type evaluation}
During subtyping, we evaluate both sides:
\begin{itemize}
  \item |A <: B|  \quad if \quad eval(|A|) |<:| eval(|B|)
\end{itemize}
\pause

Straightforward for if-then-else:
\begin{itemize}
  \item eval(|If[true, A, B]|) = |A|
  \item eval(|If[false, A, B]|) = |B|
\end{itemize}
\pause

More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
\end{slide}
\begin{slide}{Type evaluation: type tests}
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |false| \quad if |x.type| and |@\tp{T}@| are disjoint
  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |@\tp{x.isInstanceOf}@[T]| \quad otherwise
\end{itemize}
\end{slide}

@@

|foo(readInt()): { (_: Int) + 1 }|
\end{slide}

\begin{slide}{Generalized, or not singleton at all?}
\begin{code}
dependent
def int2str(x: Int)
  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ (_: String) }
  =   if (x == 0) "zero" else readString()
\end{code}
\begin{overprint}
\onslide<1>
\begin{block}{Definition}
A singleton type is a type that contains exactly one value.
\end{block}

\onslide<2>
\begin{block}{Definition}
\sout{A singleton type is a type that contains exactly one value.}
\end{block}
\end{overprint}
\end{slide}

\begin{slide}{Generalized singletons: recap}
We proposed a generalization of Scala's singleton types:
\begin{enumerate}
  \item lift a subset of Scala's language constructs to the type level
  \item add a "precise mode" of type inference
  \item evaluate types during subtyping
\end{enumerate}
\end{slide}

% % Termination
% % Virtual dispatch

@@

\begin{overprint}
\onslide<2>
Match types can be explained in terms of generalized singletons:

@@

\onslide<3>
We evaluate type tests using subtyping and type disjointness:
\begin{itemize}
  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |true@ @| \quad if |@\tp{X}@| is a subtype of |@\tp{T}@|
  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |false| \quad if |@\tp{X}@| and |@\tp{T}@| are disjoint
  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |@\tp{X.isInstanceOf}@[T]| \quad otherwise
\end{itemize}

\onslide<4>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] @=>@ elem(x.head)
  case x: Any => x
\end{code}

\onslide<5>
The type/term correspondence dictates match type semantics:

\medskip
\begin{code}
def elem[X](x: X): Elem[X] = x match
  case x: String => x.charAt(0)
  case x: List[t] @=>@ elem(x.head)
  case x: Any => x

elem[Any]("foo")
\end{code}
\vspace{-10pt}
\end{overprint}
\end{slide}

\begin{slide}{Formalization}
System FM is an extension of System \Fsub with
\begin{itemize}
  \item classes (defined externally)
  \item pattern matching
  \item match types
\end{itemize}
\pause
Design goals:
\begin{itemize}
  \item explain the essence of match types
  \item give us confidence in our design
  \item be as simple as possible
\end{itemize}
\end{slide}

@@

\begin{slide}{System FM's disjointness relation}
\begin{center}
\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
\end{center}
\end{slide}

\begin{slide}{System FM's evaluation rules}
We evaluate term-level matches by querying the class inheritance relation:
\begin{center}
\includegraphics{figures/FMEMatch.pdf}
\end{center}

\visible<2>{We evaluate type-level matches using subtyping and disjointness:}

\begin{center}
\visible<2>{\includegraphics{figures/FMSMatch.pdf}}
\end{center}
\end{slide}

@@

\begin{slide}{Empty types}

\begin{block}{Definition}
An empty type is a type that contains no values.
\end{block}

Our system does not like empty types.

@@

@@

\begin{slide}{Variance}
\begin{block}{Definitions}
|F| is covariant in |@\tp{T}@|, written |F[+T]|, if |@\tp{T1}@| |<:| |@\tp{T2}@| implies |F[T1]| |<:| |F[T2]|.\\
|G| is contravariant in |@\tp{T}@|, |G[-T]|, if |@\tp{T1}@| |<:| |@\tp{T2}@| implies |G[T1]| |>:| |G[T2]|.
\end{block}

We cannot prove disjointness when variance is involved:

$\forall$ |T1|, |T2|. \quad  |F[Nothing] <: F[T1]| \quad  and  \quad |F[Nothing] <: F[T2]|\\
$\forall$ |T1|, |T2|. \quad  |    G[Any] <: G[T1]|  \quad and  \quad |    G[Any] <: G[T2]|

@@

% \appendix

% \begin{slide}{Type-safe regular expressions (simplified)}
% \begin{code}
% val rational = new Regex("(\\d+)\\.(\\d+)")
% rational.unapply("3.1415"): Option[(String, String)]

% type GroupsOf[P <: String] =
%   Reverse[Loop[P, 0, Length[P], EmptyTuple]]

% type Loop[P, Lo, Hi, Acc <: Tuple] =
%   Lo match
%     case Hi => Acc
%     case _ => CharAt[P, Lo] match
%       case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
%       case '\\' => Loop[P, Lo + 2, Hi, Acc]
%       case _ => Loop[P, Lo + 1, Hi, Acc]
% \end{code}
% \end{slide}
